---
title: "Covariance derivative (wrt. x) under scaling (~ warping)"
author: "Y. Richet"
layout: post
date:   2018-11-22 00:00:00 +0100
categories: r jekyll
---

_"A non-stationary covariance-based Kriging method for metamodelling in engineering design"_
Y. Xiong, W. Chen, D. Apley, and X. Ding, Int. J. Numer. Meth. Engng, 2007

<hr/>


```{r, warning=FALSE}
if (!is.element("DiceKriging",installed.packages())) install.packages("DiceKriging")

# More robust (and expensive) EI optimization
if (!is.element("rgenoud",installed.packages())) install.packages("rgenoud")

# Also load DiceView for easy plot
if (!is.element("DiceOptim",installed.packages())) install.packages("DiceOptim")


library(DiceKriging)
packageDescription("DiceKriging")

# library(testthat)
```

# Covariance derivative (wrt. x)

```{r fig.height=6, fig.width=6, warning=FALSE}
f = function(x) {
    x = x^4
    1-1/2*(sin(12*x)/(1+x)+2*cos(7*x)*x^5+0.7)}
plot(f)
```




## Checking km covariance derivative 'covVector.dx'

```{r warning=FALSE, fig.height=6, fig.width=9 }
X <- matrix(c(0,.25,.5,.75,1.0),ncol=1)
y <- f(X)

set.seed(123)
m <- km(formula=~1, design=X, response=y,scaling=F,control=list(trace=FALSE))
print(m)

# plot covariance function (of x)
x = seq(0,1,,101) #101 because we need to have X values in x to check derivative is null
cx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # covariance function (in 5d space)
cdx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # derivative of covariance function
for (i in 1:length(x)) {
  mx = predict(m,x[i],type="SK",se.compute=T)
  cx[,i] = covMatrix(m@covariance,rbind(X=m@X,x[i]))$C[1:length(X),1+length(X)]
  cdx[,i] = covVector.dx(m@covariance,x=x[i],X=m@X,mx$c)
}

par(mfrow=c(2,3))
for (j in 1:length(X)) {
  plot(x, cx[j,], type='l',main=paste0('Covariance to X_',j))
  abline(v=X[j,],col='blue')
  for (i in 1:(length(x)/10))
    arrows(x[10*i], cx[j,10*i], x[10*i]+0.1, cx[j,10*i] + cdx[j,10*i]/10,length=0.05,col='red')

  ij = which(x==X[j,1])
  # testthat::test_that(desc="zero derivative at X_i",expect_true(cdx[j,ij] == 0))
}
par(mfrow=c(1,1))
```





## Checking km _scaling_ (1 node) covariance derivative 'covVector.dx'

```{r warning=FALSE, fig.height=6, fig.width=9 }
X <- matrix(c(0,.25,.5,.75,1.0),ncol=1)
y <- f(X)

set.seed(123)
m_scaling0 <- km(formula=~1, design=X, response=y,scaling=T,knots=list(design=c(.5)),control=list(trace=FALSE))
print(m_scaling0)

# plot covariance function (of x)
x = seq(0,1,,101) #101 because we need to have X values in x to check derivative is null
cx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # covariance function (in 5d space)
cdx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # derivative of covariance function
for (i in 1:length(x)) {
  mx = predict(m_scaling0,x[i],type="SK",se.compute=T)
  cx[,i] = covMatrix(m_scaling0@covariance,rbind(X=m_scaling0@X,x[i]))$C[1:length(X),1+length(X)]
  cdx[,i] = covVector.dx(m_scaling0@covariance,x=x[i],X=m_scaling0@X,mx$c)
}

par(mfrow=c(2,3))
for (j in 1:length(X)) {
  plot(x, cx[j,], type='l',main=paste0('Covariance to X_',j))
  abline(v=X[j,],col='blue')
  for (i in 1:(length(x)/10))
    arrows(x[10*i], cx[j,10*i], x[10*i]+0.1, cx[j,10*i] + cdx[j,10*i]/10,length=0.05,col='red')

  ij = which(x==X[j,1])
  # test_that(desc="zero derivative at X_i",expect_true(cdx[j,ij] == 0))
}
par(mfrow=c(1,1))
```



## Checking km _scaling_ covariance derivative 'covVector.dx'

```{r warning=FALSE, fig.height=9, fig.width=9 }
X <- matrix(c(0,0.125,.25,0.375,.5,0.625,.75,0.875,1.0),ncol=1)
y <- f(X)

set.seed(123)
m_scaling <- km(formula=~1, design=X, response=y,scaling=T,knots=list(design=c(0,.5,1)),control=list(trace=FALSE))
print(m_scaling)

# plot covariance function (of x)
x = seq(0,1,,201) #101 because we need to have X values in x to check derivative is null
cx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # covariance function (in 5d space)
cdx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # derivative of covariance function
for (i in 1:length(x)) {
  mx = predict(m_scaling,x[i],type="SK",se.compute=T)
  cx[,i] = covMatrix(m_scaling@covariance,rbind(X=m_scaling@X,x[i]))$C[1:length(X),1+length(X)]
  cdx[,i] = covVector.dx(m_scaling@covariance,x=x[i],X=m_scaling@X,mx$c)
}

par(mfrow=c(3,3))
for (j in 1:length(X)) {
  plot(x, cx[j,], type='l',main=paste0('Covariance to X_',j))
  abline(v=X[j,],col='blue')
  for (i in 1:(length(x)/10))
    arrows(x[10*i], cx[j,10*i], x[10*i]+0.1, cx[j,10*i] + cdx[j,10*i]/10,length=0.05,col='red')

  ij = which(x==X[j,1])
  # test_that(desc="zero derivative at X_i",expect_true(cdx[j,ij] == 0))
}
par(mfrow=c(1,1))
```




## Checking km _affine scaling_ (no node given) covariance derivative 'covVector.dx'

```{r warning=F, fig.height=9, fig.width=9 }
X <- matrix(c(0,0.125,.25,0.375,.5,0.625,.75,0.875,1.0),ncol=1)
y <- f(X)

set.seed(123)
m_ascaling <- km(formula=~1, design=X, response=y,scaling=T,knots=NULL,control=list(trace=FALSE))
print(m_ascaling)

# plot covariance function (of x)
x = seq(0,1,,201) #101 because we need to have X values in x to check derivative is null
cx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # covariance function (in 5d space)
cdx = matrix(NaN,nrow=nrow(X),ncol=length(x)) # derivative of covariance function
for (i in 1:length(x)) {
  mx = predict(m_ascaling,x[i],type="SK",se.compute=T)
  cx[,i] = covMatrix(m_ascaling@covariance,rbind(X=m_ascaling@X,x[i]))$C[1:length(X),1+length(X)]
  cdx[,i] = covVector.dx(m_ascaling@covariance,x=x[i],X=m_ascaling@X,mx$c)
}

par(mfrow=c(3,3))
for (j in 1:length(X)) {
  plot(x, cx[j,], type='l',main=paste0('Covariance to X_',j))
  abline(v=X[j,],col='blue')
  for (i in 1:(length(x)/10))
    arrows(x[10*i], cx[j,10*i], x[10*i]+0.1, cx[j,10*i] + cdx[j,10*i]/10,length=0.05,col='red')

  ij = which(x==X[j,1])
  # test_that(desc="zero derivative at X_i",expect_true(cdx[j,ij] == 0))
}
par(mfrow=c(1,1))
```





# ... used with EGO



```{r,echo=F, warning=F, fig.width=7, fig.height=1}
library(DiceOptim)
packageDescription("DiceOptim")
```


```{r}
EGO.nsteps = function (model, fun, nsteps, lower, upper, parinit = NULL, 
    control = NULL, kmcontrol = NULL) 
{
    n <- nrow(model@X)
    if (is.null(kmcontrol$penalty)) 
        kmcontrol$penalty <- model@penalty
    if (length(model@penalty == 0)) 
        kmcontrol$penalty <- NULL
    if (is.null(kmcontrol$optim.method)) 
        kmcontrol$optim.method <- model@optim.method
    if (is.null(kmcontrol$parinit)) 
        kmcontrol$parinit <- model@parinit
    if (is.null(kmcontrol$control)) 
        kmcontrol$control <- model@control
    control$print.level = 0
    for (i in 1:nsteps) {
        oEGO <- max_EI(model = model, lower = lower, upper = upper, 
            parinit = parinit, control = control)
        model@X <- rbind(model@X, oEGO$par)
        model@y <- rbind(model@y, fun(t(oEGO$par)))
        kmcontrol$parinit <- covparam2vect(model@covariance)
        kmcontrol$control$trace = FALSE
        kmcontrol$scaling = is.element("eta",slotNames(model@covariance))
        if (is.element("knots",slotNames(model@covariance))) kmcontrol$knots = model@covariance@knots else kmcontrol$knots = NULL
        if (model@param.estim) {
            model <- km(formula = model@trend.formula, design = model@X, 
                response = model@y, covtype = model@covariance@name, 
                lower = model@lower, upper = model@upper, nugget = NULL, 
                penalty = kmcontrol$penalty, optim.method = kmcontrol$optim.method, 
                parinit = kmcontrol$parinit, control = kmcontrol$control, 
                gr = model@gr, iso = is(model@covariance, "covIso"),
                scaling = kmcontrol$scaling, knots = kmcontrol$knots)
        }
        else {
            coef.cov <- covparam2vect(model@covariance)
            model <- km(formula = model@trend.formula, design = model@X, 
                response = model@y, covtype = model@covariance@name, 
                coef.trend = model@trend.coef, coef.cov = coef.cov, 
                coef.var = model@covariance@sd2, nugget = NULL, 
                iso = is(model@covariance, "covIso"))
        }
    }
    return(list(par = model@X[(n + 1):(n + nsteps), , drop = FALSE], 
        value = model@y[(n + 1):(n + nsteps), , drop = FALSE], 
        npoints = 1, nsteps = nsteps, lastmodel = model))
}

```




## Objective function

```{r fig.height=6, fig.width=6}
branin = function(x) {
    x = x^.5
    DiceKriging::branin(x)
}

# a 9-points factorial design, and the corresponding response
d <- 2
n <- 9
design.fact <- expand.grid(seq(0,1,length=3), seq(0,1,length=3))
names(design.fact)<-c("x1", "x2")
design.fact <- data.frame(design.fact)
names(design.fact)<-c("x1", "x2")
response.branin <- apply(design.fact, 1, branin)
response.branin <- data.frame(response.branin)
names(response.branin) <- "y"

.x = seq(0,1,,51)
.p3d = persp(.x,.x,matrix(apply(expand.grid(.x,.x),1,branin),ncol=length(.x)),zlab = "branin(sqrt(.))", phi = 60,theta = 30)
points(trans3d(design.fact[,1],design.fact[,2],response.branin$y,.p3d),col='black',pch=20) 
```

## Basic (no scaling)

```{r fig.height=6, fig.width=6 }
set.seed(123)

# model identification
fitted.model1 <- km(~1, design=design.fact, response=response.branin,
covtype="gauss", control=list(pop.size=50,trace=FALSE), parinit=c(0.5, 0.5))

# EGO n steps
library(rgenoud)
nsteps <- 10 
lower <- rep(0,d)
upper <- rep(1,d)
oEGO <- EGO.nsteps(model=fitted.model1, fun=branin, nsteps=nsteps,
lower=lower, upper=upper, control=list(pop.size=20, BFGSburnin=2))
print(oEGO$par)
print(oEGO$value)

.p3d = persp(.x,.x,matrix(apply(expand.grid(.x,.x),1,branin),ncol=length(.x)),zlab = "branin(sqrt(.))", phi = 60,theta = 30)
points(trans3d(oEGO$lastmodel@X[,1],oEGO$lastmodel@X[,2],apply(oEGO$lastmodel@X,1,branin),.p3d),col='black',pch=20) 
points(trans3d(oEGO$par[,1],oEGO$par[,2],apply(oEGO$par,1,branin),.p3d),col='red',pch=20) 
```


## With scaling

```{r fig.height=6, fig.width=6 }
set.seed(123)

# model identification
fitted.model1 <- km(~1, design=design.fact, response=response.branin,
covtype="gauss", control=list(pop.size=50,trace=FALSE), parinit=c(0.5, 0.5), scaling = T)

# EGO n steps
library(rgenoud)
nsteps <- 10
lower <- rep(0,d)
upper <- rep(1,d)
oEGO <- EGO.nsteps(model=fitted.model1, fun=branin, nsteps=nsteps,
lower=lower, upper=upper, control=list(pop.size=20, BFGSburnin=2))
print(oEGO$par)
print(oEGO$value)

.p3d = persp(.x,.x,matrix(apply(expand.grid(.x,.x),1,branin),ncol=length(.x)),zlab = "branin(sqrt(.))", phi = 60,theta = 30)
points(trans3d(oEGO$lastmodel@X[,1],oEGO$lastmodel@X[,2],apply(oEGO$lastmodel@X,1,branin),.p3d),col='black',pch=20) 
points(trans3d(oEGO$par[,1],oEGO$par[,2],apply(oEGO$par,1,branin),.p3d),col='red',pch=20) 
```
