---
title: "Optimization for (simulation) engineering"
author: "Y. Richet"
layout: post
date:   2018-04-25 00:00:00 +0100
categories: r jekyll
---

## Optimization for (simulation) engineering

Engineering objective: optimize $f$ function/simulator, with lowest $f$ evaluations as possible.

<center>

```{r,echo=F, warning=F, fig.height=4, fig.width=7}
#f=function(x) 1-sin(.25+pi*x^.75)
f=(function(x) 1-sin(-.20+3*pi*x^.85)+x/3+5*(x-.45)^2)
#x0 = c(0.05,.5,.9)
x0 = c(0.0, 0.3, 0.7, 1.0)
y0 = f(x0)
xx = seq(0,1,,100)

#plot(f)
#points(x0,y0)

grey = rgb(.75,.75,.75)
plot_f = function(xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F,lty=2,xy0=T,...) {
    par(mar=c(1,1,1,1))
    par(bg=NA)
    
    plot(0,0,type='n',xlim=c(0,1),ylim=c(min(f(xx)),max(f(xx))),xaxt=xaxt, yaxt=yaxt, ann=ann,axes=axes,frame.plot=frame.plot,...)
    rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)
    if (length(xx)!= length(f(xx))) {print(xx)
    print(f(xx))}
    lines(xx,f(xx),lty=lty,col='white',lwd=2,xlab="x",ylab="f(x)")
    if (xy0) points(x0,y0,col='red',pch=20)
}
plot_f(lty=1,xy0=F)
i_min_f = which.min(f(xx))
points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
```
</center>



## Basic idea

create some $\color{blue}{models\ of\ f}$ based on few evaluations $x={X}$

<center>
```{r,echo=F, warning=F, fig.height=4, fig.width=7}
plot_f()
i_min_f = which.min(f(xx))
points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')

if (!is.element("DiceKriging",installed.packages())) install.packages("DiceKriging")
library(DiceKriging)
k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
make_km = function(n=10){
    mx=matrix(NaN,n,length(xx))
    for (i in 1:n) {
        set.seed(i)
        mx[i,] = simulate(k,nsim=1,newdata=xx,cond=T,nugget.sim=0.0000001)
    }
    mx
}

mx = make_km()
for (i in 1:nrow(mx)) {
    lines(xx,mx[i,],col='blue')
    points(x0,y0,col='blue')

    i_min = which.min(mx[i,])
    points(xx[i_min],mx[i,i_min],col='blue',pch='x')
}
```
</center>


## (simple) Kriging

$$\color{blue}M(x) = \mathcal{N}(m(x),s^2(x))$$


* $m(x) = C(x)^T C(X)^{-1} f(X)$
* $s^2(x) = c(x) - C(x)^T C(X)^{-1} C(x)$
* $C$ is the covariance kernel  $C(.) = C(X,.)$, $c(.) = C(x,.)$


<center>
```{r,echo=F, warning=F, fig.height=4, fig.width=7}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')

points(x0,y0,col='blue')
k_x = predict.km(k,xx,"UK")
for (p in c(3,2,1))
    polygon(c(xx,rev(xx)),c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),col=rgb(0,0,1,.2),border=NA)
```
</center>


## Efficient Global Optimization


<center>
```{r,echo=F, warning=F, fig.width=4, fig.height=3}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
</center>

</div>

Let's define the _Expected Improvement_:

$$\color{violet}{EI}(x) = E[(min \{ f(X) \} - M(x))^+]$$

_which is (also) analytical thanks to $M$ properties..._



$$u(x) = {{min\{f(X)\} - m(x)} \over {s(x)}}$$


<center>
```{r,echo=F, warning=F, fig.width=4, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)
```
</center>


$$\color{violet}{EI}(x) = s(x)\ \big(\ u(x)p_{\mathcal{N}}(u(x)) + d_{\mathcal{N}}(u(x))\ \big)$$


_"Efficient Global Optimization of Expensive Black-Box Functions"_- Jones, Schonlau, Welch,  
(Journal of Global Optimization, December 1998)

EGO:  
Maximize $EI(x)$ (*), compute $f$ there, add to $X$.  
Repeat until ...

<hr/>

* __+__ good trade-off between exploration and exploitation
* __+__ requires few evaluations of $f$
* __-__ often lead to add close points to each others ...  
Which is not very comfortable for kriging numerical stability
* __-__ "one step lookahead" (myopic) strategy
* __-__ rely on model suitability to $f$

(*) using standard optimization algorithm: BFGS, PSO, DiRect, ...


## EGO - step 0

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>


## EGO - step 1

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>


## EGO - step 2

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>



## EGO - step 3

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>



## EGO - step 4

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>



## EGO - step 5

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>


## EGO - step 6

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>


## EGO - step 7

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>


## EGO - step 8

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>


## EGO - step 9

<center>
```{r,echo=F, warning=F, fig.width=7, fig.height=4}
plot_f()
#i_min_f = which.min(f(xx))
#points(xx[i_min_f],f(xx[i_min_f]),col='red',pch='x')
abline(h=min(y0),col='red')

k=km(design = data.frame(x=x0),response = data.frame(y=y0),coef.cov = 0.2,coef.var = .2,control=list(trace=F))
k_x = predict.km(k,xx,"UK")

points(x0,y0,col='blue')
for (p in c(3,2,1)) {
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),max,min(y0)),col=rgb(0,0,1,.2),border=NA)
    polygon(c(xx,rev(xx)),lapply(c(k_x$mean+k_x$sd*p,rev(k_x$mean-k_x$sd*p)),min,min(y0)),col=rgb(.8,0,.8,.2),border=NA)
}
```
```{r,echo=F, warning=F, fig.width=7, fig.height=1}
par(mar=c(0,1,0,1))

uxx = (min(y0)-k_x$mean)/k_x$sd
EI_xx = k_x$sd * (uxx * pnorm(uxx) + dnorm(uxx))
EI_xx[is.nan(EI_xx)]=0

plot(0,0,type='n',xlim=c(0,1),ylim=c(0,max(EI_xx,na.rm = T)),xaxt='n', yaxt='n', ann=FALSE,axes=F,frame.plot=F)
rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], border=NA, col = grey
)

polygon(xx,EI_xx, col=rgb(1,0,1,.5), border=NA)

x0 = c(x0, xx[which(EI_xx==max(EI_xx,na.rm=T))[1]])
y0 = f(x0)
```
</center>



